Open Data, Open Source and Open Standards in chemistry - The Blue Obelisk five years on

### 2011-05-12 paper.tex
Tidy up section on 2nd gen tools

Although feature-rich and robust cheminformatics toolkits are useful
in and of themselves, they can also be seen as providing a base layer
on which additional tools and applications can be built. This is one
of the reasons that cheminformatics toolkits are so important to the
open source `ecosystem'; their availability lowers the barrier for the
development of a `second generation' of chemistry software that no
longer needs to concern itself with the low-level details of
manipulating chemical structures, and can focus on providing
additional functionality and ease-of-use.

Bioclipse~\cite{Spjuth:2007fk} (v2.4 released in Aug 2010) and Avogadro
\cite{WebAvogadro} (v1.0 in Oct 2009) are two examples of such software, based
on the CDK and Open Babel, respectively. Bioclipse is an award-winning
molecular workbench for life sciences
which wraps cheminformatics functionality behind user-friendly interfaces and
graphical editors while Avogadro is a 3D molecular editor and viewer aimed at
preparing and analysing computational chemistry calculations. Both
projects have are designed to be extended or scripted by users through
a plugin architective and the provision of scripting support (using
Bioclipse Scripting Language~\cite{Bioclipse2}, or Python in the case
of Avogadro).
An interesting
aspect of both Avogadro and Bioclipse is that they share some developers
with the underlying toolkits and this has driven the development of new
features in the CDK and Open Babel.

@@

CDK-Taverna \cite{Kuhn:2010p4001} as an open-source cheminformatics workflow solution.
Commonly used in this context are workflow engines for cheminformatics, where numerous recurring tasks can be automated, including tasks for chemical data filtering, transformation, curation and migration workflows, chemical documentation and information retrieval related workflows (structures, reactions, pharmacophores, object relational data etc.) or data analysis workflows (statistics and clustering/ machine learning for QSAR, diversity analysis etc.
CDK-Taverna has matured to become a freely available and increasingly powerful tool for the biosciences. It was recently ported to
Taverna version 2.x.
The combination of CDK-Taverna workers with the multitude of workers
already published on myexperiment.org by the large and active user
community of Taverna enables scientists to quickly build workflows to
process a diverse set of data as typically found in today's systems
biology scenarios.

Some other recent projects that build on Blue Obelisk software include
AMBIT (a GUI that facilitates registration of chemicals for the REACH
EU directive on toxicity, based on the CDK), etc.

@@

>>>>

Although feature-rich and robust cheminformatics toolkits are useful
in and of themselves, they can also be seen as providing a base layer
on which additional tools and applications can be built. This is one
of the reasons that cheminformatics toolkits are so important to the
open source `ecosystem'; their availability lowers the barrier for the
development of a `second generation' of chemistry software that no
longer needs to concern itself with the low-level details of
manipulating chemical structures, and can focus on providing
additional functionality and ease-of-use. Although a wide range of
chemistry software has been built using Blue Obelisk
components (see for example, the ``Related Software'' link on the Open
Babel website, or ``Software using CDK'' at the CDK website), in this
section we focus on second-generation tools which themselves have been
developed by members of the Blue Obelisk.

Bioclipse~\cite{Spjuth:2007fk} (v2.4 released in Aug 2010) and Avogadro
\cite{WebAvogadro} (v1.0 in Oct 2009) are two examples of such software, based
on the CDK and Open Babel, respectively. Bioclipse is an award-winning
molecular workbench for life sciences
that wraps cheminformatics functionality behind user-friendly interfaces and
graphical editors while Avogadro is a 3D molecular editor and viewer aimed at
preparing and analysing computational chemistry calculations. Both
projects are designed to be extended or scripted by users through
the provision of a plugin architective and scripting support (using
Bioclipse Scripting Language~\cite{Bioclipse2}, or Python in the case
of Avogadro).
An interesting
aspect of both Avogadro and Bioclipse is that they share some developers
with the underlying toolkits and this has driven the development of new
features in the CDK and Open Babel.

@@

A final example of 2nd-generation Blue Obelisk software is the 
AMBIT2~\cite{WebAMBIT} software, a GUI that facilitates registration
of chemicals for the REACH EU directive on toxicity, and which is
based on the CDK.

@@

Another way through which interoperability of Blue Obelisk projects
has been promoted and developed is through integration into
workflow software such as Taverna~\cite{Hull:2006p60} and
KNIME~\cite{WebKNIME} (both open source).
Such software makes it easy to automate recurring
tasks, and to combine analyses or data from a variety of different software
and web services.
A combination of the Chemistry Development Kit and Taverna, for instance, was
reported in 2010~\cite{Kuhn:2010p4001}. 
In the case of KNIME, it comes with built-in basic collection of CDK-based and
Open Babel-based nodes, while other nodes for the RDKit and Indigo are
available from KNIME's ``Community Updates'' site.
